Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-22 ms.Rnw
minor edits

While this work is focused on the models proposed in \cite{lee20}, we believe that our suggestions are relevant to a larger class of problems. 
In order to determine the extent to which this work generalizes, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, and model.
The search resulted in 66 papers, of which 32 proposed or used dynamic models to describe the cholera epidemic in Haiti.
This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling cholera in Haiti.
% Although the literature review focused on cholera and Haiti, 
The wide relevance of our findings in these papers---supplemented with studies of other epidemiological and dynamic systems---suggests that this work may also be relevant to a much broader class of problems.

@@

Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).

In the following subsections, complete descriptions of Models~1--3 are provided.
While the model description is scientifically critical, as well as being necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.

@@

We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}) and forecasting (Sec.~\ref{sec:filter}).

@@

All three model considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
In the following subsections we describe our approach to fitting these mechanistic models.

@@

One approach to modeling a dynamic system is through probabilistic models.
With a probabilistic model, we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
Using a probabilistic model results in several advantages, including the ability to account for variability present in the system.
% which is of great interest to scientists and policy makers.
Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.
The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.

There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
% including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
\end{equation}
Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
% The inclusion of a possible trend in transmission is due to observing an apparent continual decrease in reported cholera infections from 2012-2019 can be seen in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.

@@

Model~1 is implemented using the \code{pomp} package \cite{pomp2}, which contains an implementation of the IF2 algorithm that was used to compute the likelihood profile.
% Performing a profile confidence search of a parameter also results in a MLE calculation, as the value that maximizes the profile likelihood also maximizes the entire likelihood surface \citep{murphy2000}.
Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.

@@

Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or SpatPOMP) with no randomness in the dynamic process or measurement.
By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.

\cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.

The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years epidemic.

@@

\subsubsection{Model~3}
Model~3 is also of a probabilistic model.
In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
% We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.

@@

Parameters that must be fit in Model~3 are primarily shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by fitting independent department-level models to the data.
The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
A first order approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
\begin{equation}
  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
\end{equation}
This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.

@@

We fit the PanelPOMP version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function.
Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.

@@

Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package, similar to what was done in \cite{lee20}.
Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.

@@

% While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, which is model specific and beyond the scope of this article.

In this case study, a brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
Consider the unexpectedly large exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
Whatever the case may be, this fitting procedure results in model likelihoods unable to improve upon a log-linear benchmark (see Sec.~\ref{sec:model_diagnostics} and Tab.~\ref{tab:likes}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons provide insight into how each model performs relative to the others.
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmarks}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
The benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation, and comparing any statistic of the observed data to simulations from the model---which is sometimes referred to as diagnostic probes.

@@

The central goal of a forecast is to provide an accurate estimate of the future state of a system based on currently available data.
When a mechanistic model is used, forecasts may also provide estimates of the future effects of potential interventions.
Forecasting models are built using available scientific understanding, but forecasting can also be a way of testing new scientific hypotheses in real time \citep{lewis22}.
In order to provide trustworthy information, however, the reliability of the forecast should be undersood.
In particular, researchers should account for various forms of uncertainty present in model forecasts, and  calibrate the proposed model to observed data.

>>>>

While this work is focused on the models proposed in \cite{lee20}, we believe that our suggestions are relevant to a larger class of problems.
In order to determine the extent to which this work generalizes, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, and model.
The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling an infectious disease outbreak.
% Although the literature review focused on cholera and Haiti,
The wide relevance of our findings in these papers---supplemented with studies of other epidemiological and dynamic systems---suggests that this work may also be relevant to a much broader class of problems.

@@

Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model $\modelCounter$.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).

In the following subsections, complete descriptions of Models~1--3 are provided.
While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
A first-time reader may choose to skim through the rest of this section, and return later.

@@

We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and interpreting model fit (Sec.~\ref{sec:science}).

@@

Each of the three models considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
In the following subsections we describe our approach to fitting these mechanistic models.

@@

One approach to modeling a dynamic system is through probabilistic models.
With such a model, the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$ is supposed, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
Under this framework, the observed data $y_{\seq{1}{N}}^*$---along with some unobserved value $\bm{X}^{(\modelCounter)}_{\seq{0}{N}} = x^*_{\seq{0}{N}}$---are assumed to be a single realization of the model $\big(x^*_{\seq{0}{N}}, y_{\seq{1}{N}}^*\big) \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
This modeling framework results in several advantages, including the ability to account for variability present in the system.
Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.
% The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.

Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.

The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
\end{equation}
Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.

@@

Model~1 is implemented using the \code{pomp} package \cite{pomp2}, relying heavily on the source code provided by \cite{lee20}, who also used the MIF2 algorithm to estimate $\paramVec$.
Aside from the introduction of the parameter $\transmissionTrend$ and the multiplicative Gamma-distributed white noise $d\Gamma(t)/dt$ in the process model (Eq.~\ref{model1:lambda}), the only change made that could increase the likelihood of the model was using a substantially increased computational effort to maximize the likelihood.
While flexible, plug-and-play methods like MIF2 require a large number of computations, and the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles used and the number of MIF iterations.
The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.

% Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.

@@

Model~2 is a deterministic model, which are degenerate probabilistic models that contain no randomness in the dynamic process or measurement.
By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.

\cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
% A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.

@@

\subsubsection{Model~3}
Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
The decision to model a dynamic system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.

@@

Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the MIF2 algorithm to separately estimate parameters for independent department-level models;
the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
A first order approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
\begin{equation}
  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
\end{equation}
This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.

@@

We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function.
Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.

@@

Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.

@@

% While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, the answer to which is model specific and beyond the scope of this case study.

A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
Consider the unexpectedly large exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
Whatever the case may be, this fitting procedure results in model likelihoods unable to improve upon a log-linear benchmark (see Sec.~\ref{sec:model_diagnostics} and Tab.~\ref{tab:likes}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.

@@

Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
These comparisons provide insight into how each model performs relative to the others.
To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
The benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.

@@

Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing any statistic of the observed data to simulations from the model \citep{wood10}---which is sometimes referred to as diagnostic probes \citep{king16}.

@@

The central goal of a forecast is to provide an accurate estimate of the future state of a system based on currently available data.
When a mechanistic model is used, forecasts may also provide estimates of the future effects of potential interventions.
Forecasting models are built using available scientific understanding, but forecasting can also be a way of testing new scientific hypotheses in real time \citep{lewis22}.
In order to provide trustworthy information, however, the reliability of the forecast should be understood.
In particular, researchers should account for various forms of uncertainty present in model forecasts, and  calibrate the proposed model to observed data.

The most recently available information about a dynamic system should be used to obtain reliable forecasts.
Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$, where $\bm{x}^{(\modelCounter)}_N$ is the value of the latent state at time $N$.
Because the latent state $\bm{X}^{(\modelCounter)}_N$ is unobservable, draws from the desired conditional density are unobtainable.
Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
Obtaining forecasts in this fashion allows us to obtain results that are consistent with the model assumptions and the observed data.
In this particular case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing.
This alone partially explains the unrealistic forecasts of \cite{lee20}, which were obtained by simulating the model from initial conditions;
Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
% \eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
